Simple repression draft

### 2020-05-16 doc/paper.tex
Edits from RP & HG

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
Part of its utility in dissecting transcriptional regulation is
its ratiometric nature, which is always easier than making an
absolute measurement. Also, by measuring otherwise identical cells
with and without a transcription factor present, any biological
noise common to both conditions can be made to cancel away.

@@

\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
What have we learned from our exploration of the models
in~\fig{fig1:means_cartoons} and their predictions of
fold-changes in mean expression levels? The takeaway is the
existence of a master curve (\fig{fig1:means_cartoons}C) to which
the fold-change predictions of all the models collapse. This
master curve is parametrized by only two effective parameters:
$\Delta F_R$, which characterizes the repressor, and $\rho$,
which characterizes all other features of the promoter
architecture. The key assumption underpinning this result is that
zero transcription occurs when a repressor is bound to its
operator, though note we are agnostic about the molecular
mechanism which achieves this.\footnote{
Steric effects are the obvious mechanism, but, for instance,
``action-at-a-distance'' mediated by kinked DNA due to
repressors bound tens or hundreds of nucleotides upstream of a
promoter is plausible as well.}

@@

As suggested in~\cite{Chure2019}, we believe this master curve
should generalize to architectures with multiple repressor
binding sites, as long as the exclusivity of transcription factor
binding and transcription initation is maintained. The
interpretation of $\Delta F_R$ is then of an effective free
energy of all repressor bound states.
In an equilibrium picture this is simply given by the log of the
sum of Boltzmann weights of all repressor bound states, which
looks like the log of a partition function of a subsystem.
In a nonequilibrium picture, while we can still mathematically gather
terms and give the resulting collection the label $\Delta F_R$,
it is unclear if the physical interpretation as an effective free energy
makes sense, and we leave this an open problem for future work.\footnote{
As is well known, free energies cannot be assigned unambiguously
to states out of equilibrium because the free energy change along
a generic path traversing the state space is path dependent,
unlike at equilibrium.
A consequence of this is that, out of equilibrium, $\Delta F_R$
is no longer a simple sum of Boltzmann weights. Instead it
resembles a restricted sum of King-Altman
diagrams~\cite{King1956, Hill1966}. Following the work of
Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
as an effective free energy, but this remains unclear to us.
}

@@

One may worry that our ``one curve to rule them all'' is a
mathematical tautology. In fact we \textit{agree} with this
criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
be meaningfully interpreted as a real, physical free energy.
We believe our theory deserves to be scrapped unless it can be
demonstrated that $\Delta F_R$ \textit{is} a physically
meaningful and interpretable free energy.

How do we test this and probe the theory with fold-change
measurements? There is a fundamental limitation in that the
master curve is essentially a one-parameter function of $\Delta
F_R + \log\rho$. Worse, there are many \textit{a priori}
plausible microscopic mechanisms that could contribute to the
value of $\rho$, such as RNAP binding and escape
kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
more speculatively, RNAP clusters analogous to those obvserved in
eukaryotes~\cite{Cisse2013}. Even if $\Delta F_R$ is measured to
high precision, inferring the potential microscopic contributions
to $\rho$, buried inside a log no less, from fold-change
measurements seems hopeless. As a statistical inference problem
it is entirely nonidentifiable, in the language
of~\cite{Gelman2013}, Section 4.3.

If we cannot simply infer values of $\rho$ from measurements of
fold-change, can we perturb some of the parameters that make up
$\rho$ and measure the change? Unfortunately this is probably
off-limits experimentally: most of the potential contributors to
$\rho$ are global processes that affect many or all genes. For
instance, changing RNAP association rates by changing RNAP copy
numbers, or changing supercoiling kinetics by changing
topoisomerase copy numbers, would massively perturb the entire
cell's physiology and confound any determination of $\rho$.

@@

In light of this, we prefer the view that parameters and rates
are not directly comparable between cartoons
in~\fig{fig1:means_cartoons}. Rather, parameters in the
simpler cartoons represent coarse-grained combinations of
parameters in the finer-grained models. For instance, by equating
$\rho$ between any two models, one can derive various possible
correspondences between the two models' parameters. Note that
these correspondences are clearly not unique, since many possible
associations could be made. It then is a choice as to what microscopic
interpretations the model-builder prefers for the parameters in a
particular cartoon, and as to which coarse-grainings lend
intuition and which seem nonsensical. Indeed, since it remains an open question what microscopic features dominate $\rho$ (as suggested above, perhaps RNAP binding and escape
kinetics~\cite{DeHaseth1998, Mitarai2015}, or supercoiling
accumulation and release~\cite{Chong2014, Sevier2016}, or,
something more exotic like RNAP clusters~\cite{Cisse2013}), we are hesitant to put too much weight on any one microscopic interpretation of model parameters that make up $\rho$.

@@

The realization that $\rho\approx1$ to an excellent
approximation, \textit{independent} of which model
in~\ref{fig1:means_cartoons} one prefers, goes a long way towards
explaining the surprising success of equilibrium models of simple
repression. Even though our 2- and 3-state models get so many
details of transcription wrong, it does not matter because
fold-change is a cleverly designed ratio. Since $\rho$ subsumes
all details except the repressor, and $\log\rho\approx0$, fitting
these simple models to fold-change measurements can still give a
surprisingly good estimate of repressor binding energies. So the
ratiometric construction of fold-change fulfills its intended
purpose of canceling out all features of the promoter
architecture except the repressor itself. Nevertheless it is
perhaps surprising how effectively it does so: \textit{a priori},
one might not have expected $\rho$ to be quite so close to 1.

>>>>

Fold-change has proven a very convenient observable in past
work~\cite{Garcia2011a, Razo-Mejia2018, Chure2019}.
Part of its utility in dissecting transcriptional regulation is
its ratiometric nature, which is always easier than making an
absolute measurement. Also, by measuring otherwise identical cells
with and without a transcription factor present, any biological
noise common to both conditions can be made to cancel away.
In addition, the presence of different operator sites is known to
change the basal expression level in the absence of
repressor~\mmnote{look up cite}.
Computing fold-changes neatly cancels this effect as well.

@@

\subsubsection{Summary and discussion of results across models for fold-changes in mean expression}
What have we learned from our exploration of the models
in~\fig{fig1:means_cartoons} and their predictions of
fold-changes in mean expression levels? The takeaway is the
existence of a master curve (\fig{fig1:means_cartoons}C) to which
the fold-change predictions of all the models collapse. This
master curve is parametrized by only two effective parameters:
$\Delta F_R$, which characterizes the repressor, and $\rho$,
which characterizes all other features of the promoter
architecture. The key assumption underpinning this result is that
zero transcription occurs when a repressor is bound to its
operator. Note, however, that we are agnostic about the molecular
mechanism which achieves this;
steric effects are the obvious mechanism, but, for instance,
``action-at-a-distance'' mediated by kinked DNA due to
repressors bound tens or hundreds of nucleotides upstream of a
promoter is plausible as well.

@@

As suggested in~\cite{Chure2019}, we believe this master curve
should generalize to architectures with multiple repressor
binding sites, as long as the exclusivity of transcription factor
binding and transcription initation is maintained. The
interpretation of $\Delta F_R$ is then of an effective free
energy of all repressor bound states.
In an equilibrium picture this is simply given by the log of the
sum of Boltzmann weights of all repressor bound states, which
looks like the log of a partition function of a subsystem.
In a nonequilibrium picture, while we can still mathematically gather
terms and give the resulting collection the label $\Delta F_R$,
it is unclear if the physical interpretation as an effective free energy
makes sense.
As is well known, free energies cannot be assigned unambiguously
to states out of equilibrium because the free energy change along
a generic path traversing the state space is path dependent,
unlike at equilibrium.
A consequence of this is that, out of equilibrium, $\Delta F_R$
is no longer a simple sum of Boltzmann weights. Instead it
resembles a restricted sum of King-Altman
diagrams~\cite{King1956, Hill1966}. Following the work of
Hill~\cite{Hill1989}, it may yet be possible to interpret this expression
as an effective free energy, but this remains unclear to us.
We leave this an open problem for future work.

@@

One may worry that our ``one curve to rule them all'' is a
mathematical tautology. In fact we \textit{agree} with this
criticism if $\Delta F_R$ is ``just a fitting parameter'' and cannot
be meaningfully interpreted as a real, physical free energy.
An analogy to Hill functions is appropriate here. One of their
great strengths and weaknesses, depending on the use they are put
to, is that their parameters coarse-grain many details and are
generally not interpretable in terms of microscopic models, for
deep reasons discussed at length in~\cite{Frank2013}.
By contrast, our master curve claims to have the best of both worlds:
a coarse-graining of all details besides the repressor into
a single effective parameter $\rho$,
while simultaneously retaining an interpretation of
$\Delta F_R$ as a physically
meaningful and interpretable free energy.
Our task, then, is to prove or disprove this claim.

How do we test this and probe the theory with fold-change
measurements? There is a fundamental limitation in that the
master curve is essentially a one-parameter function of $\Delta
F_R + \log\rho$. Worse, there are many \textit{a priori}
plausible microscopic mechanisms that could contribute to the
value of $\rho$, such as RNAP binding and escape
kinetics~\cite{DeHaseth1998, Mitarai2015}, and/or supercoiling
accumulation and release~\cite{Chong2014, Sevier2016}, and/or,
RNAP clusters analogous to those obvserved in
eukaryotes~\cite{Cisse2013, Cho2016}
and recently also observed in bacteria~\cite{Ladouceur2020}.
Even if $\Delta F_R$ is measured to
high precision, inferring the potential microscopic contributions
to $\rho$, buried inside a log no less, from fold-change
measurements seems hopeless. As a statistical inference problem
it is entirely nonidentifiable, in the language
of~\cite{Gelman2013}, Section 4.3.

If we cannot simply infer values of $\rho$ from measurements of
fold-change, can we perturb some of the parameters that make up
$\rho$ and measure the change? Unfortunately we suspect this is
off-limits experimentally: most of the potential contributors to
$\rho$ are global processes that affect many or all genes. For
instance, changing RNAP association rates by changing RNAP copy
numbers, or changing supercoiling kinetics by changing
topoisomerase copy numbers, would massively perturb the entire
cell's physiology and confound any determination of $\rho$.

@@

In light of this, we prefer the view that parameters and rates
are not directly comparable between cartoons
in~\fig{fig1:means_cartoons}. Rather, parameters in the
simpler cartoons represent coarse-grained combinations of
parameters in the finer-grained models. For instance, by equating
$\rho$ between any two models, one can derive various possible
correspondences between the two models' parameters. Note that
these correspondences are clearly not unique, since many possible
associations could be made. It then is a choice as to what microscopic
interpretations the model-builder prefers for the parameters in a
particular cartoon, and as to which coarse-grainings lend
intuition and which seem nonsensical.
Indeed, since it remains an open question what microscopic
features dominate $\rho$ (as suggested above, perhaps RNAP
binding and escape kinetics~\cite{DeHaseth1998, Mitarai2015}, or
supercoiling accumulation and release~\cite{Chong2014, Sevier2016},
or, something more exotic like
RNAP clusters~\cite{Cisse2013, Cho2016, Ladouceur2020}),
we are hesitant to put too much weight on any one microscopic
interpretation of model parameters that make up $\rho$.

@@

The realization that $\rho\approx1$ to an excellent
approximation, \textit{independent} of which model
in~\fig{fig1:means_cartoons} one prefers, goes a long way towards
explaining the surprising success of equilibrium models of simple
repression. Even though our 2- and 3-state models get so many
details of transcription wrong, it does not matter because
fold-change is a cleverly designed ratio. Since $\rho$ subsumes
all details except the repressor, and $\log\rho\approx0$, fitting
these simple models to fold-change measurements can still give a
surprisingly good estimate of repressor binding energies. So the
ratiometric construction of fold-change fulfills its intended
purpose of canceling out all features of the promoter
architecture except the repressor itself. Nevertheless it is
perhaps surprising how effectively it does so: \textit{a priori},
one might not have expected $\rho$ to be quite so close to 1.
